Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community

### 2020-10-21 manuscript.rmd
Corrected variable symbols from italic to roman

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

@@

## Model for effective rooting depth (ERD)
\
To estimate effective rooting depth, we tested alternate models for growth (Eqs. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance ($K_{leaf}$) approximated using rhizosphere soil water potential ($\Psi_{soil}$) as a proxy for the soil water limiting factor on growth. In our models $K_{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

For each species, we estimated diurnal maximum value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

@@

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

@@

#### Leaf hydraulic vulnerability curves
\
We obtained leaf hydraulic vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

@@

As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the 31 ERD species we estimated species $\Psi_{50,~leaf}$ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

@@

#### Leaf Area Index (LAI)
\
To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless). For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq. \@ref(eq:M-chosen).

@@

### Model validation and selection
\
To validate the ERD models we used data for xylem sap $\delta^2H$ (permil) as an independent observation of ERD. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species xylem sap $\delta^2H$ (permil). We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig. S\@ref(fig:iso-data)). We removed species that are likely to have no leaves in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. We were thus left with 7 species with data on $\delta^2H$ as well as modelled estimates of ERD.

## ERD coordination with above-ground hydraulic traits
\
To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$). See [SI](#SIData.brett) for data collection and estimation of these variables.

@@

We regressed species ERD against species ($Mort_{mean}$). To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$. $\Psi_{crit}$ was defined as $\Psi_{50,~leaf}$.

@@

  Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. S\@ref(fig:swp-droughts)). Nevertheless, across the simulation period (1990-2015) $\Psi_{soil,~z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

@@

## Leaf hydraulic vulnerability curves
\
For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs. $\Psi_{leaf}$, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); *Adj. R*^2^ = `r afit.r2`; *p* `r  afit.p`; Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r bfit.r2`; *p* `r  bfit.p`; also see Fig. S\@ref(fig:obs-plc-curves). Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.  S\@ref(fig:obs-comm-plc-curves)). 
Fig. \@ref(fig:ab-fit) here.

## Effective rooting depths (ERD)
\
ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig. \@ref(fig:erd-iso)). The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit. ERD from all the other models showed a positive, but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq. \@ref(eq:M-chosen)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

## Effective rooting depths, experienced drought and mortality
\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig. S\@ref(fig:erd-mort-census)). Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi_{ERD}$ < $\Psi_{crit}$ (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety. Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD. We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.

\
Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *p* = 0.075), and other causes may explain observed mortality [@McDowell.2018].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests. Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.

\
We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). We find that leaf conductance declined more rapidly with declining $\Psi_{leaf}$ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

@@

Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from model based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *p*-values are given in inset.

Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta^2H$ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \@ref(tab:ab-param) for the full form of scientific names.

>>>>

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

@@

## Model for effective rooting depth (ERD)
\
To estimate effective rooting depth, we tested alternate models for growth (Eqs. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance (K~leaf~) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth. In our models K~leaf~ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

For each species, we estimated diurnal maximum value of $K~{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi$~soil,~i~|~z,~h~ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

@@

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi$~soil,z~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

@@

#### Leaf hydraulic vulnerability curves
\
We obtained leaf hydraulic vulnerability curves (K~leaf~ vs. $\Psi$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

@@

As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the 31 ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

@@

#### Leaf Area Index (LAI)
\
To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless). For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq. \@ref(eq:M-chosen).

@@

### Model validation and selection
\
To validate the ERD models we used data for xylem sap $\delta$^2H (permil) as an independent observation of ERD. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species xylem sap $\delta$^2H (permil). We used xylem sap $\delta$^2H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta$^2H in the soil (Fig. S\@ref(fig:iso-data)). We removed species that are likely to have no leaves in March as xylem sap $\delta$^2H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. We were thus left with 7 species with data on $\delta$^2H as well as modelled estimates of ERD.

## ERD coordination with above-ground hydraulic traits
\
To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance (K~max,stem~), leaf turgor loss point ($\Psi$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi$~88,stem~), and hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~). See [SI](#SIData.brett) for data collection and estimation of these variables.

@@

We regressed species ERD against species ($Mort_{mean}$). To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi$~soil,z~=~ERD~) were below species $\Psi$~crit~. $\Psi$~crit~ was defined as $\Psi$~50,leaf~.

@@

  Soil Water Potential dynamics ($\Psi$~soil,z~) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. S\@ref(fig:swp-droughts)). Nevertheless, across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

@@

## Leaf hydraulic vulnerability curves
\
For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for K~leaf~ vs. $\Psi$~leaf~, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); *Adj. R*^2^ = `r afit.r2`; *p* `r  afit.p`; Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r bfit.r2`; *p* `r  bfit.p`; also see Fig. S\@ref(fig:obs-plc-curves). Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.  S\@ref(fig:obs-comm-plc-curves)). 
Fig. \@ref(fig:ab-fit) here.

## Effective rooting depths (ERD)
\
ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2H), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig. \@ref(fig:erd-iso)). The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit. ERD from all the other models showed a positive, but non-significant relationship with $\delta$^2H at $\alpha$ = 0.05 (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq. \@ref(eq:M-chosen)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2H values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

## Effective rooting depths, experienced drought and mortality
\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig. S\@ref(fig:erd-mort-census)). Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi$~ERD~ was more negative than species $\Psi$~crit~ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi$~ERD~ < $\Psi$~crit~ (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety. Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015). Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z~=~ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD. We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

\
Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *p* = 0.075), and other causes may explain observed mortality [@McDowell.2018].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests. Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta$^2H.

\
We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

@@

Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs. $\\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from model based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *p*-values are given in inset.

Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta$^2H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \@ref(tab:ab-param) for the full form of scientific names.
